The effect of detector nonlinearity on \wfa\ PSF profiles for weak gravitational lensing measurements

### 2016-04-19 bias_from_NL.tex

An ideal detector would produce a measurable signal that is proportional to the detected photons. However, there are several places in the signal chain where this expected linearity is not realized, and the conversion of charge to measured voltage (or digital numbers) becomes non-linear. First of all, the charge accumulation rate might be a function of the photon-accumulation rate. This type of effect is known as reciprocity failure (RF), and it is a flux-dependent type of nonlinearity (\citealt{smith08}, \citealt{biesiadzinski11}). In addition to RF, the p-n junction acts as a parallel-plate capacitor, and as charge accumulates the depletion region narrows, causing a deviation from linearity of the charge-to-voltage conversion relation. Nonlinearity can also be introduced through the electronic gain of the ROIC. 
The last two types of nonlinearity depend on fluence (total flux per pixel, or integrated signal, as opposed to RF, which is flux dependent), and can be analyzed together in a singe transfer function typically called ``nonlinearity" (NL). We study the impacts of NL (more relevant at high signals) on PSF measurements in this paper, while we leave investigations on the consequences of RF (relevant at lower signals) on WL measurements for future work.

%\textcolor{blue}{Following \citealt{hilbert04}, \citealt{hilbert08}, and \cite{hilbert14} (which study non-linearity in the Hubble Space Telescope's Wide Field Camera (WFC3) NIR detector\textemdash a H1RG 1.7 $\mu$m cutoff detector, similar to the H4Rg that will be used in WFIRST), we model the measured voltage $S$ corresponding to the number of electron-hole pairs $Q$ as a fourth-order polynomial of the form:}
To obtain a functional model of NL as a function of signal, we use laboratory measurements\footnote{Performed at Caltech by R. Smith and collaborators.} of H2RG 1.7 $\mu$m cutoff detectors, which have the same basic design of the H4RG that will be used in the WFI but with 18 $\mu$m pixels. A constant flux source is illuminated on the detector, and the NL measurements are done by non-destructive sample-up-the-ramp (SUTR) readout at multiple times. The value of the first frame is subtracted, and then the \emph{mean} signal is fitted to a polynomial. These measurements show that the voltage relation for a measured signal $S$ is well fit by a quadratic function in the signal $Q$ of the form:
\begin{align}
S(Q)= Q + \beta Q ^2  (\mathrm{e^-}) 
\label{NL}
\end{align}
%\begin{align}
%\hat{p} = \frac{\frac{\partial I}{\partial p} C^{-1} (I - I_{p=0})} {\frac{\partial I}{\partial p} C^{-1} \frac{\partial I}{\partial p}} \\
%\end{align}
%To determine the scale of spatial variation in NL among the individual pixels, high S/N co-added flats were taken at different exposure times and constant flux, and their ratios taken (after subtraction of a mean dark flat for each frame). The sequence of means of the ratio images followed the quadratic function of Eq. \ref{NL}, with a standard variation that includes shot noise and deviations of the individual NL curves from the mean NL. After subtracting in quadrature the contribution from shot noise, a remaining floor in the variation of about $12\%$ r.m.s. can be attributed to the variation of the NL coefficient $\beta$.  
%\textcolor{blue}{For the pixels in the WFC3 NIR detector, \citealt{hilbert14} calculate that $\beta \in [-0.5,1.5]\times10^{-6}$, $\gamma \in [-1,0.5]\times10^{-10}$, and $\delta \in [-1,2]\times10^{-15}$. Therefore, in our calculations we use the midpoints these intervals as nominal values for each parameter, \emph{i.e.} $(\beta_0$, $\gamma_0$, $\delta_0$) = ($0.5\times10^{-6}$, $-0.25\times10^{-10}$, $1.5\times10^{-15}$). In addition, we will assume that the coefficients in each pixel can be drawn from a Gaussian distribution with means $(\beta_0$, $\gamma_0$, $\delta_0$) and ($\sigma_{\beta}$, $\sigma_{\gamma}$, $\sigma_{\delta}$) = ($5.9\times10^{-7}$, $5.625\times10^{-11}$, $1.35\times10^{-15}$)  (dispersion values taken as the mean of the standard deviations calculated in four cuadrants of the WFC3 H1RG detector by \citealt{hilbert08})}
Thus, in our calculations we assume that each individual pixel satisfies a NL function of the same form as Eq. \ref{NL}. In addition, when we study the impact of spatial variation of the model coefficient along the pixel array we will assume that the parameter $\beta$ in each pixel can be drawn from a Normal distribution of the form $\mathcal{N} \sim (\beta_0,\ \sigma_{\beta_0})$, where the nominal mean value $\beta_0$ is measured to be $-3.566\times10^{-7}$/e$^{-}$. This mean nominal value implies an attenuation of the signal of about $4\%$ for a typical pixel full well depth in the WFI detectors of $1.1 \times 10^5$ electrons (Eq. \ref{NL}).  
For this work, we use profiles with a simulated AB magnitude up to 18.3 for an exposure time of 168.1 seconds (expected for the HLS). The total flux of the profiles in each band is shown in Table 1, along with their peak value. When the profile is drawn on a pixelated postage stamp with a scale of $p=0.11$ arcseconds and placed on the center of the pixel\footnote{Note that the peak value is a function of the PSF profile centroid location within the pixel, as well as resolution. When drawing the PSF profiles in our simulations, we randomize the PSF centroid within the native scale pixel, as described in Section YYY.}, this particular magnitude produces a peak charge of $\sim 1e5$e$^{-}$ (Y106 band, see Table 1), which represents about 90\% of the typical pixel full well value assumed above.
%which represents a total flux (over all pixels of the postage stamp where each profile was drawn) of up to $\sim 2.82 \times 10^5 \ \mathrm{e^-}$ (J129 band, see Table 1), with a resulting in a signal attenuation of about $3.4\%$.  
Calculations with the {\tt{Trilegal}} galaxy model\footnote{\url{http://stev.oapd.inaf.it/cgi-bin/trilegal}} \citep{girardi12} show that there will be approximately 20 stars at or brighter than this magnitude and flux level per detector.\footnote{We thank Christopher Hirata, who performed these calculations. The NIR filter sets used in {\tt{Trilegal}} had to be converted from the vega to the AB magnitude system. The code was run at the South Galactic Pole with the SDSS and 2MASS ($ugriz$ and $JHK_\mathrm{s}$) filters in 1 deg$^2$ and interpolated to WFIRST filter centers, which ignores the fact that stars have spectral structure in the NIR, but gives a result good to about $10\%$. The number of stars quoted in the text is expected to be higher at moderate galactic latitudes.} 
% (Be aware that many of the NIR filter sets in Trilegal are on the Vega system and have to be converted to AB.) I did a run at the South Galactic Pole with the SDSS + 2MASS (ugriz + JHKs) filters in 1 deg^2 and interpolated to WFIRST filter centers. [This isn't really a great procedure since the stars have spectral structure in the NIR, but at the "tens of percents" level -- which I think is what you're asking for -- this should be fine.] In 1 deg^2, I get 964 saturated stars in J, 943 in H, and 667 in F184. This corresponds to 15.0, 14.7, and 10.4 saturated stars per detector. At moderate Galactic latitude, these numbers will be higher. Also each detector will have a software-defined guide window that will be placed on one of the stars that would saturate in the full-length exposure.}  }  
We assume that galaxies will have about two orders of magnitude fewer total electrons than bright stars, and therefore for this quadratic model, star shapes are distorted by NL and galaxies are (approximately) not, which would result in an incorrect PSF correction if not accurately calibrated. We apply this mapping to simulated \wfa\ PSF profiles and quantify the impact on PSF properties such as size and shape. 
%We have verified that the results using the mean NL coefficient for all pixels do not differ significatively from those using the distribution above for each pixel (see Section and Table 2 below). 

>>>>

An ideal detector would produce a measurable signal that is proportional to the detected photons. However, there are several places in the signal chain where this expected linearity is not realized, and the conversion of charge to measured voltage (or digital numbers) becomes non-linear. First of all, the charge accumulation rate might be a function of the photon-accumulation rate. This type of effect is known as reciprocity failure (RF), and it is a flux-dependent type of nonlinearity (\citealt{smith08}, \citealt{biesiadzinski11}). In addition to RF, each pixel's p-n junction acts as a parallel-plate capacitor, and as charge accumulates the depletion region narrows, causing a deviation from linearity of the charge-to-voltage conversion relation. Nonlinearity can also be introduced through the electronic gain of the ROIC. 
The last two types of nonlinearity depend on fluence (total flux per pixel, or integrated signal, as opposed to RF, which is flux dependent), and can be analyzed together in a singe transfer function typically called ``nonlinearity" (NL). We study the impacts of NL (more relevant at high signals) on PSF measurements in this paper, while we leave investigations on the consequences of RF (relevant at lower signals) on WL measurements for future work.

Assuming that the dominant contribution to nonlinearity is the varying capacitance of the pixel p-n junction, the correction to the detector signal is well approximated by a quadratic term:
\begin{align}
S(Q)= Q + \beta Q ^2 % (\mathrm{e^-})
%\textcolor{blue}{Following \citealt{hilbert04}, \citealt{hilbert08}, and \cite{hilbert14} (which study non-linearity in the Hubble Space Telescope's Wide Field Camera (WFC3) NIR detector\textemdash a H1RG 1.7 $\mu$m cutoff detector, similar to the H4Rg that will be used in WFIRST), we model the measured voltage $S$ corresponding to the number of electron-hole pairs $Q$ as a fourth-order polynomial of the form:}
%To obtain a functional model of NL as a function of signal, we use laboratory measurements\footnote{Performed at Caltech by R. Smith and collaborators.} of H2RG 1.7 $\mu$m cutoff detectors, which have the same basic design of the H4RG that will be used in the WFI but with 18 $\mu$m pixels. A constant flux source is illuminated on the detector, and the NL measurements are done by non-destructive sample-up-the-ramp (SUTR) readout at multiple times. The value of the first frame is subtracted, and then the \emph{mean} signal is fitted to a polynomial. These measurements show that the voltage relation for a measured signal $S$ is well fit by a quadratic function in the signal $Q$ of the form:
%\begin{align}
%S(Q)= Q + \beta Q ^2  (\mathrm{e^-}) 
\label{NL}
\end{align}
Here, $Q$ is the true number of elementary charges collected in the pixel, $S$ is the number inferred from the voltage change at the sense node, and $\beta$ is a constant.  To estimate $\beta$, we compute $Q(V)=C(V)*V/q_e$, where the total capacitance is given by the varying junction capacitance plus a constant $C(V) = C_{\rm jn}(V) + C_{\rm fix}$. \textbf{[CITE McCAUGHREAN]}  The junction capacitance varies as
\begin{equation}
C_{\rm jn}(V) \sim (1+V/V_{bi})^{-1/2}
\end{equation}
Here, $V_{bi}$ is the ``built in'' potential of the junction and $V=V_{\rm DSUB}-V_{\rm RESET}+\delta V$, where $V_{\rm DSUB}$ is the constant potential at the diode cathode, $V_{\rm RESET}$ is the initial potential at the anode, and $\delta V$ is the change in anode potential due to accumulated photocharge.  As an example, we plug in measurements of a 2.4$\mu$m cutoff H2RG by Finger (2006): \textbf{[CITE FINGER]}% https://www.eso.org/sci/meetings/2006/neon-2006/Finger.pdf
\begin{eqnarray*}
V_{\rm bi} &=& 0.412{\mbox V} \\
V_{\rm DSUB} &=& 1{\mbox V} \\
V_{\rm RESET} &=& 0.5{\mbox V} \\
C_{\rm fix} &=& 17.8 {\mbox fF} \\
C_{\rm jn}(V=.912{\mbox V}) &=& 30 {\mbox fF}
\end{eqnarray*}¥
Using these parameters to compute $Q(V)$ for $ 0\le \delta V \le 0.3$ V (corresponding to maximum $Q$=60115) and then inverting to find $S(Q)$, we find it is well-fit (to 0.1\% or better) by (\ref{NL}) with $\beta$=-1.18e-6.  In practice, HxRG calibrations have included additional polynomial parameters to reduce residuals, extend the range of valid $Q$, and account for additional nonlinear effects \textbf[CITE HILBERT PAPERS]; however, the additional parameters are highly degenerate, resulting in large variances in the fitted values.  For our purposes, it suffices to analyze the shape-distorting effects of nonlinearity using a single parameter, $\beta$, which encapsulates most of the effect.

%\textcolor{blue}{Following \citealt{hilbert04}, \citealt{hilbert08}, and \cite{hilbert14} (which study non-linearity in the Hubble Space Telescope's Wide Field Camera (WFC3) NIR detector\textemdash a H1RG 1.7 $\mu$m cutoff detector, similar to the H4Rg that will be used in WFIRST), we model the measured voltage $S$ corresponding to the number of electron-hole pairs $Q$ as a fourth-order polynomial of the form:}
%To obtain a functional model of NL as a function of signal, we use laboratory measurements\footnote{Performed at Caltech by R. Smith and collaborators.} of H2RG 1.7 $\mu$m cutoff detectors, which have the same basic design of the H4RG that will be used in the WFI but with 18 $\mu$m pixels. A constant flux source is illuminated on the detector, and the NL measurements are done by non-destructive sample-up-the-ramp (SUTR) readout at multiple times. The value of the first frame is subtracted, and then the \emph{mean} signal is fitted to a polynomial. These measurements show that the voltage relation for a measured signal $S$ is well fit by a quadratic function in the mean signal $\langle Q \rangle$ of the form:
%\begin{align}
%\hat{p} = \frac{\frac{\partial I}{\partial p} C^{-1} (I - I_{p=0})} {\frac{\partial I}{\partial p} C^{-1} \frac{\partial I}{\partial p}} \\
%\end{align}
%To determine the scale of spatial variation in NL among the individual pixels, high S/N co-added flats were taken at different exposure times and constant flux, and their ratios taken (after subtraction of a mean dark flat for each frame). The sequence of means of the ratio images followed the quadratic function of Eq. \ref{NL}, with a standard variation that includes shot noise and deviations of the individual NL curves from the mean NL. After subtracting in quadrature the contribution from shot noise, a remaining floor in the variation of about $12\%$ r.m.s. can be attributed to the variation of the NL coefficient $\beta$.  
%\textcolor{blue}{For the pixels in the WFC3 NIR detector, \citealt{hilbert14} calculate that $\beta \in [-0.5,1.5]\times10^{-6}$, $\gamma \in [-1,0.5]\times10^{-10}$, and $\delta \in [-1,2]\times10^{-15}$. Therefore, in our calculations we use the midpoints these intervals as nominal values for each parameter, \emph{i.e.} $(\beta_0$, $\gamma_0$, $\delta_0$) = ($0.5\times10^{-6}$, $-0.25\times10^{-10}$, $1.5\times10^{-15}$). In addition, we will assume that the coefficients in each pixel can be drawn from a Gaussian distribution with means $(\beta_0$, $\gamma_0$, $\delta_0$) and ($\sigma_{\beta}$, $\sigma_{\gamma}$, $\sigma_{\delta}$) = ($5.9\times10^{-7}$, $5.625\times10^{-11}$, $1.35\times10^{-15}$)  (dispersion values taken as the mean of the standard deviations calculated in four cuadrants of the WFC3 H1RG detector by \citealt{hilbert08})}

%Thus, in our calculations we assume that each individual pixel satisfies a NL function of the same form as Eq. \ref{NL}. In addition, when we study the impact of spatial variation of the model coefficient along the pixel array we will assume that the parameter $\beta$ in each pixel can be drawn from a Normal distribution of the form $\mathcal{N} \sim (\beta_0,\ \sigma_{\beta_0})$, where the nominal mean value $\beta_0$ is measured to be $-3.566\times10^{-7}$/e$^{-}$.

%This mean nominal value implies an attenuation of the signal of about $4\%$ for 
For this work, we assume a typical pixel full well depth in the WFI detectors of $1.1 \times 10^5$ electrons (Eq. \ref{NL}).  
Using PSF profiles with a simulated AB magnitude up to 18.3 for an exposure time of 168.1 seconds (expected for the HLS), the total flux in each band is shown in Table 1, along with the peak pixel value. When the profile is drawn on a pixelated postage stamp with a scale of $p=0.11$ arcseconds and placed on the center of the pixel\footnote{Note that the peak value is a function of the PSF profile centroid location within the pixel, as well as resolution. When drawing the PSF profiles in our simulations, we randomize the PSF centroid within the native scale pixel, as described in Section YYY.}, this particular magnitude produces a peak charge of $\sim 1e5$e$^{-}$ (Y106 band, see Table 1), which represents about 90\% of the typical pixel full well value assumed above.
\textbf{[CALCULATE NL \% for nominal beta]}
%which represents a total flux (over all pixels of the postage stamp where each profile was drawn) of up to $\sim 2.82 \times 10^5 \ \mathrm{e^-}$ (J129 band, see Table 1), with a resulting in a signal attenuation of about $3.4\%$.  
Calculations with the {\tt{Trilegal}} galaxy model\footnote{\url{http://stev.oapd.inaf.it/cgi-bin/trilegal}} \citep{girardi12} show that there will be approximately 20 stars at or brighter than this magnitude and flux level per detector.\footnote{We thank Christopher Hirata, who performed these calculations. The NIR filter sets used in {\tt{Trilegal}} had to be converted from the vega to the AB magnitude system. The code was run at the South Galactic Pole with the SDSS and 2MASS ($ugriz$ and $JHK_\mathrm{s}$) filters in 1 deg$^2$ and interpolated to WFIRST filter centers, which ignores the fact that stars have spectral structure in the NIR, but gives a result good to about $10\%$. The number of stars quoted in the text is expected to be higher at moderate galactic latitudes.} 
% (Be aware that many of the NIR filter sets in Trilegal are on the Vega system and have to be converted to AB.) I did a run at the South Galactic Pole with the SDSS + 2MASS (ugriz + JHKs) filters in 1 deg^2 and interpolated to WFIRST filter centers. [This isn't really a great procedure since the stars have spectral structure in the NIR, but at the "tens of percents" level -- which I think is what you're asking for -- this should be fine.] In 1 deg^2, I get 964 saturated stars in J, 943 in H, and 667 in F184. This corresponds to 15.0, 14.7, and 10.4 saturated stars per detector. At moderate Galactic latitude, these numbers will be higher. Also each detector will have a software-defined guide window that will be placed on one of the stars that would saturate in the full-length exposure.}  }  
We assume that galaxies will have about two orders of magnitude fewer total electrons than bright stars, and therefore for this quadratic model, star shapes are distorted by NL and galaxies are (approximately) not, which would result in an incorrect PSF correction if not accurately calibrated. We apply this mapping to simulated \wfa\ PSF profiles and quantify the impact on PSF properties such as size and shape. 
%We have verified that the results using the mean NL coefficient for all pixels do not differ significatively from those using the distribution above for each pixel (see Section and Table 2 below). 
% DC offsets which are intrinsic to the readout circuit and which vary from pixel to pixel
